---
title: "Understanding the Factors <br> Influencing Seoul Bike Sharing Demand"
subtitle: "IBM 6540: Final Report Presentation"
author: "**Yujin Jeong**"

format:
  revealjs:
    theme: simple
    slide-number: true
    transition: fade
    incremental: false
    width: 1280
    height: 720
    margin: 0.06
    controls: true
    progress: true
    center: true
execute:
  echo: false
  warning: false
  message: false
include-in-header:
  text: |
    <style>
    :root {
      --bg: #effcf8;
      --mint: #00796b;
      --mint2: #2aa99a;
      --pink: #f8a8c4;
      --pink-dark: #d96f9c;
      --blue: #7ec8e3;
      --text: #264653;
    }
    html, body, .reveal, .reveal .slides {
      background: var(--bg) !important;
    }
    .reveal {
      font-size: 25px;
      color: var(--text);
    }
    .reveal .slides section {
      text-align: left;
    }
    .slide-content {
      width: 86%;
      max-width: 1080px;
      margin-left: auto;
      margin-right: auto;
      box-sizing: border-box;
    }
    .slide-content ul,
    .slide-content ol {
      display: block !important;
      margin-left: 0 !important;
      margin-right: 0 !important;
      padding-left: 1.25em !important;
      box-sizing: border-box;
    }
    .slide-content li {
      margin-bottom: 0.28em;
      line-height: 1.32;
    }
    .reveal h1 {
      color: var(--mint);
      font-weight: 760;
      letter-spacing: -0.02em;
    }
    .reveal h2 {
      color: var(--mint);
      font-weight: 730;
      border-bottom: 5px solid var(--pink);
      padding-bottom: 0.12em;
      margin-bottom: 0.45em;
    }
    .reveal h3 {
      color: var(--blue);
      font-weight: 700;
      margin-bottom: 0.25em;
    }
    .reveal strong {
      color: var(--pink-dark);
      font-weight: 760;
    }
    .reveal a { color: var(--mint); }
    .reveal table {
      font-size: 0.58em;
      width: 100%;
      border-collapse: collapse;
    }
    .reveal th {
      color: var(--mint);
      border-bottom: 2px solid var(--pink);
    }
    .reveal td, .reveal th { padding: 0.25em 0.35em; }
    .plot-img {
      display: block;
      max-height: 470px;
      max-width: 90%;
      margin: 0.25em auto 0 auto !important;
      border-radius: 14px;
      box-shadow: 0 8px 20px rgba(0, 121, 107, 0.16);
    }
    .note {
      color: #3e5f68;
      font-size: 0.82em;
      margin-top: 0.6em;
    }
    .pill {
      display: inline-block;
      background: #ffffff99;
      border: 2px solid var(--pink);
      border-radius: 999px;
      padding: 0.18em 0.65em;
      margin: 0.12em 0.1em;
      color: var(--mint);
      font-weight: 700;
    }
    .kpi-grid {
      display: grid;
      grid-template-columns: repeat(4, 1fr);
      gap: 0.7em;
      margin-top: 0.5em;
    }
    .kpi {
      background: rgba(255, 255, 255, 0.68);
      border-left: 6px solid var(--mint2);
      border-radius: 14px;
      padding: 0.7em;
      box-shadow: 0 5px 16px rgba(0, 121, 107, 0.08);
    }
    .kpi .num {
      font-size: 1.35em;
      color: var(--mint);
      font-weight: 800;
    }
    .kpi .label {
      font-size: 0.75em;
      color: var(--text);
    }
    .two-col {
      display: grid;
      grid-template-columns: 1fr 1fr;
      gap: 1.2em;
      align-items: start;
    }
    .compact { font-size: 0.82em; }
    .center-title { text-align: center; }
    </style>
---

```{r setup}
# Data preparation used throughout the report
bike_raw <- read.csv("SeoulBikeData.csv", fileEncoding = "latin1", check.names = FALSE)

# Rename columns to analysis-friendly names
names(bike_raw) <- c(
  "date", "rented_bike_count", "hour", "temperature", "humidity",
  "wind_speed", "visibility", "dew_point", "solar_radiation",
  "rainfall", "snowfall", "season", "holiday", "functioning_day"
)

bike_raw$date <- as.Date(bike_raw$date, format = "%d/%m/%Y")
bike_raw$weekday <- weekdays(bike_raw$date)
bike_raw$is_weekend <- ifelse(bike_raw$weekday %in% c("Saturday", "Sunday"), 1, 0)
bike_raw$rush_hour <- ifelse(bike_raw$hour %in% c(7, 8, 9, 17, 18, 19), 1, 0)
bike_raw$rain_status <- ifelse(bike_raw$rainfall > 0, "Rain", "No rain")

bike <- subset(bike_raw, functioning_day == "Yes")
bike$season <- factor(bike$season, levels = c("Spring", "Summer", "Autumn", "Winter"))
bike$holiday <- factor(bike$holiday)

mint <- "#00796b"
pink <- "#d96f9c"
light_pink <- "#f8a8c4"
blue <- "#7ec8e3"
bg <- "#effcf8"
text_col <- "#264653"

plot_theme <- function() {
  par(bg = bg, col.axis = text_col, col.lab = text_col, col.main = mint,
      mar = c(4.2, 4.5, 3.2, 1.2), family = "sans")
}
```

::: {.slide-content .center-title}
# Understanding the Factors <br> Influencing Seoul Bike Sharing Demand

### Cover Slide:

[Seoul Bike Sharing]{.pill} [Demand Analytics]{.pill} [Operations Management]{.pill}
:::

## Table of Contents

::: slide-content
1.  Introduction
2.  Population and Data Collection
3.  Data Wrangling and Final Usable Sample
4.  Sample Characteristics
5.  Research Question 1
6.  Variables, Scales, Visualization, and Modeling
7.  Summary Findings and Business Implications
:::

## Introduction: Problem Definition

::: slide-content
Bike-sharing systems need to place the **right number of bikes** at the **right time**.

-   Underestimating demand creates bike shortages and customer dissatisfaction.
-   Overestimating demand wastes operational resources.
-   Demand changes by hour, weather, season, holidays, and system availability.

### Managerial problem

Operations managers need a data-driven way to forecast hourly bike rental demand.
:::

## Importance of the Research

::: slide-content
Understanding bike demand helps managers improve:

-   Bicycle allocation across stations and operating periods
-   Staffing and maintenance planning
-   Service reliability during peak demand
-   Cost control during low-demand periods
-   Customer satisfaction through better availability
:::

## Research Objective and RQ1

::: slide-content
### Research Objective

Identify the weather and time-related factors that influence hourly bike rental demand in Seoul.

### RQ1

**Which weather and time-related factors have the greatest influence on <br> hourly bike rental demand in Seoul?**

### H1

Weather conditions and time-related variables significantly affect bike rental demand.
:::

## Population

::: slide-content
The population is all hourly bike rental observations recorded by the Seoul Bike Sharing system during the study period.

**Each observation represents one hour and includes:**

-   Rented bike count
-   Date and hour
-   Weather conditions
-   Season and holiday indicators
-   System functioning status
:::

## Data Collection

:::::::::::::::: slide-content
### Source

Seoul Bike Sharing Demand Dataset from the UCI Machine Learning Repository.

### Data type

Secondary observational data collected from a bike-sharing system and matched with weather and calendar information.

### Raw dataset

::::::::::::::: kpi-grid
::::: kpi
::: num
8,760
:::

::: label
Hourly observations
:::
:::::

::::: kpi
::: num
14
:::

::: label
Original variables
:::
:::::

::::: kpi
::: num
0
:::

::: label
Missing values
:::
:::::

::::: kpi
::: num
0
:::

::: label
Duplicate rows
:::
:::::
:::::::::::::::
::::::::::::::::

## From Records to Data

::: slide-content
The raw data is already in a tidy record format:

-   Each row = one hourly observation
-   Each column = one measured variable
-   The dependent variable is **Rented Bike Count**
-   Independent variables include weather, time, season, holiday, and functioning day variables
:::

## Data Wrangling Process

::: slide-content
The data preparation process included:

1.  Renaming variables for easier analysis
2.  Converting the Date field to date format
3.  Checking missing values and duplicates
4.  Creating weekday and weekend indicators
5.  Creating rush-hour and peak-hour indicators
6.  Removing non-functioning system hours from the modeling sample
:::

## Final Usable Sample Decision

:::::::::::::::: slide-content
The dataset includes hours when the bike-sharing system was not functioning.

-   Non-functioning hours do not represent normal customer demand.
-   These observations show constrained demand because users could not rent bikes.
-   They were removed for the RQ1 demand model.

::::::::::::::: kpi-grid
::::: kpi
::: num
8,760
:::

::: label
Raw observations
:::
:::::

::::: kpi
::: num
295
:::

::: label
Non-functioning hours removed
:::
:::::

::::: kpi
::: num
8,465
:::

::: label
Final usable sample
:::
:::::

::::: kpi
::: num
96.6%
:::

::: label
Retained for analysis
:::
:::::
:::::::::::::::
::::::::::::::::

## Sample Characteristics

:::::::::::::::: slide-content
After removing non-functioning hours, the final usable sample has the following demand characteristics:

::::::::::::::: kpi-grid
::::: kpi
::: num
729
:::

::: label
Mean hourly rentals
:::
:::::

::::: kpi
::: num
542
:::

::: label
Median hourly rentals
:::
:::::

::::: kpi
::: num
2
:::

::: label
Minimum hourly rentals
:::
:::::

::::: kpi
::: num
3,556
:::

::: label
Maximum hourly rentals
:::
:::::
:::::::::::::::

### Interpretation

Demand is highly variable, meaning operations managers need flexible planning rather than a fixed allocation rule.
::::::::::::::::

## Seasonal Demand Pattern

:::: slide-content
```{r avg_by_season, fig.width=8.5, fig.height=4.8}
plot_theme()
season_mean <- tapply(bike$rented_bike_count, bike$season, mean, na.rm = TRUE)
barplot(season_mean,
        col = c(blue, mint, light_pink, "#b9e8df"),
        border = NA,
        main = "Average Bike Rentals by Season",
        xlab = "Season",
        ylab = "Average rentals")
text(x = seq_along(season_mean) * 1.2 - 0.5,
     y = season_mean + 30,
     labels = round(season_mean, 0),
     col = text_col)
```

::: note
Summer and autumn show the highest average demand, while winter is the lowest-demand season.
:::
::::

## Hourly Demand Pattern

:::: slide-content
```{r avg_by_hour, fig.width=8.5, fig.height=4.8}
plot_theme()
hour_mean <- tapply(bike$rented_bike_count, bike$hour, mean, na.rm = TRUE)
plot(as.numeric(names(hour_mean)), hour_mean,
     type = "b", pch = 19, lwd = 3, col = mint,
     main = "Average Bike Rentals by Hour",
     xlab = "Hour of Day",
     ylab = "Average rentals")
grid(col = "#b7dcd5")
```

::: note
Demand peaks in the evening commuting period, especially around 6 PM.
:::
::::

## Variables and Their Roles

::::: {.slide-content .two-col}
<div>

### Dependent Variable

-   **Rented Bike Count**

### Weather IVs

-   Temperature
-   Humidity
-   Wind speed
-   Visibility
-   Dew point temperature
-   Solar radiation
-   Rainfall
-   Snowfall

</div>

<div>

### Time and Calendar IVs

-   Hour
-   Season
-   Holiday
-   Weekend indicator
-   Rush-hour indicator

### Control Decision

-   Functioning Day used to define the final usable sample

</div>
:::::

## Operationalization of the Scale

::: {.slide-content .compact}
| Variable              | Role |    Scale | Operational Definition           |
|-----------------------|-----:|---------:|----------------------------------|
| Rented Bike Count     |   DV |    Ratio | Number of bikes rented per hour  |
| Temperature           |   IV | Interval | Degrees Celsius                  |
| Dew Point Temperature |   IV | Interval | Degrees Celsius                  |
| Humidity              |   IV |    Ratio | Percent humidity                 |
| Wind Speed            |   IV |    Ratio | Meters per second                |
| Visibility            |   IV |    Ratio | Visibility measured in 10m units |
| Solar Radiation       |   IV |    Ratio | MJ/m²                            |
| Rainfall              |   IV |    Ratio | Millimeters per hour             |
| Snowfall              |   IV |    Ratio | Centimeters per hour             |
| Hour                  |   IV |    Ratio | Hour of day, 0-23                |
| Season                |   IV |  Nominal | Spring, Summer, Autumn, Winter   |
| Holiday               |   IV |  Nominal | Holiday vs. No Holiday           |
:::

## Data Visualization Plan for RQ1

::: slide-content
The visual analysis focuses on patterns that directly answer RQ1:

-   Hourly demand line chart
-   Seasonal demand comparison
-   Temperature vs. demand scatter plot
-   Rain vs. no-rain demand comparison
-   Correlation heatmap for numerical predictors
-   Feature importance chart from the predictive model
:::

## Correlation Overview

:::: slide-content
```{r correlation_heatmap, fig.width=8, fig.height=6, out.width="65%"}
plot_theme()
num_vars <- bike[, c("rented_bike_count", "hour", "temperature", "humidity",
                    "wind_speed", "visibility", "dew_point", "solar_radiation",
                    "rainfall", "snowfall")]
cm <- cor(num_vars, use = "complete.obs")
labels <- c("Demand", "Hour", "Temp", "Humidity", "Wind", "Visibility", "Dew Point", "Solar", "Rain", "Snow")
image(1:ncol(cm), 1:nrow(cm), t(cm[nrow(cm):1, ]),
      axes = FALSE,
      col = colorRampPalette(c("#f8a8c4", "white", "#00796b"))(100),
      main = "Correlation Heatmap")
axis(1, at = 1:ncol(cm), labels = labels, las = 2, cex.axis = 0.75)
axis(2, at = 1:nrow(cm), labels = rev(labels), las = 2, cex.axis = 0.75)
for (i in 1:nrow(cm)) {
  for (j in 1:ncol(cm)) {
    text(j, nrow(cm) - i + 1, sprintf("%.2f", cm[i, j]), cex = 0.65, col = text_col)
  }
}
box(col = mint)
```

::: note
Temperature, hour, dew point, solar radiation = positively related to demand <br> Humidity, rainfall, snowfall = negatively related to demand
:::
::::

## Temperature and Demand

:::: slide-content
```{r temperature_demand, fig.width=8.5, fig.height=4.8}
plot_theme()
plot(bike$temperature, bike$rented_bike_count,
     pch = 16, cex = 0.45, col = rgb(0, 121, 107, 80, maxColorValue = 255),
     main = "Demand Increases with Temperature",
     xlab = "Temperature (C)",
     ylab = "Rented Bike Count")
fit <- loess(rented_bike_count ~ temperature, data = bike)
temp_grid <- seq(min(bike$temperature), max(bike$temperature), length.out = 150)
lines(temp_grid, predict(fit, newdata = data.frame(temperature = temp_grid)),
      col = pink, lwd = 4)
```

::: note
Demand generally rises with temperature, although extremely high or low demand also depends on hour and season.
:::
::::

## Rainfall and Demand

:::: slide-content
```{r rainfall_demand, fig.width=7.2, fig.height=4.8}
plot_theme()
boxplot(rented_bike_count ~ rain_status,
        data = bike,
        col = c(blue, light_pink),
        border = mint,
        main = "Rainy Hours Have Lower Demand",
        xlab = "Rain Status",
        ylab = "Rented Bike Count")
```

::: note
Rainy hours have visibly lower demand, supporting the expectation that poor weather reduces bike usage.
:::
::::

## Modeling Approach

::: slide-content
Two models were used to evaluate RQ1:

1.  **Multiple Linear Regression**
    -   Interpretable baseline model
    -   Useful for understanding linear relationships
2.  **Random Forest Regression**
    -   Captures nonlinear patterns and interactions
    -   Useful for identifying the strongest predictors

### Model evaluation metrics

RMSE, MAE, and R² were used to compare model performance.
:::

## Model Performance

::: slide-content
| Model                      |  RMSE |   MAE |    R² |
|----------------------------|------:|------:|------:|
| Multiple Linear Regression | 395.1 | 302.8 | 0.602 |
| Random Forest Regression   | 187.7 | 107.6 | 0.910 |

### Interpretation

The Random Forest model performed substantially better, suggesting that bike demand is affected by nonlinear and interacting factors.
:::

## Feature Importance

:::: slide-content
```{r feature_importance, fig.width=8.5, fig.height=5}
plot_theme()
features <- c("hour", "temperature", "humidity", "wind_speed", "visibility",
              "dew_point", "solar_radiation", "rainfall", "snowfall",
              "is_weekend", "rush_hour")

if (requireNamespace("randomForest", quietly = TRUE)) {
  set.seed(42)
  rf <- randomForest::randomForest(
    rented_bike_count ~ hour + temperature + humidity + wind_speed + visibility +
      dew_point + solar_radiation + rainfall + snowfall + season + holiday +
      is_weekend + rush_hour,
    data = bike,
    ntree = 300,
    importance = TRUE
  )
  imp_raw <- randomForest::importance(rf)[, 1]
  imp <- as.numeric(imp_raw)
  names(imp) <- rownames(randomForest::importance(rf))
} else {
  imp_raw <- cor(bike[, features], bike$rented_bike_count, use = "complete.obs")
  imp <- abs(as.numeric(imp_raw[, 1]))
  names(imp) <- rownames(imp_raw)
}

imp <- imp[order(imp, decreasing = TRUE)]
imp <- imp[seq_len(min(10, length(imp)))]
friendly <- tools::toTitleCase(gsub("_", " ", names(imp)))

par(mar = c(4.2, 9.5, 3.2, 1.2), bg = bg, col.axis = text_col, col.lab = text_col, col.main = mint)
barplot(rev(as.numeric(imp)), horiz = TRUE, names.arg = rev(friendly), las = 1,
        col = mint, border = NA,
        main = "Top Predictors of Bike Demand",
        xlab = "Relative importance")
```

::: note
Temperature, hour, and dew point are the strongest predictors.
:::
::::

## Prediction Quality

:::: slide-content
```{r actual_predicted, fig.width=6.8, fig.height=5.2}
plot_theme()
set.seed(42)
train_id <- sample(seq_len(nrow(bike)), size = floor(0.8 * nrow(bike)))
train <- bike[train_id, ]
test <- bike[-train_id, ]

if (requireNamespace("randomForest", quietly = TRUE)) {
  model <- randomForest::randomForest(
    rented_bike_count ~ hour + temperature + humidity + wind_speed + visibility +
      dew_point + solar_radiation + rainfall + snowfall + season + holiday +
      is_weekend + rush_hour,
    data = train,
    ntree = 300
  )
  pred <- predict(model, newdata = test)
  title_text <- "Random Forest: Actual vs. Predicted"
} else {
  model <- lm(rented_bike_count ~ hour + temperature + humidity + wind_speed + visibility +
                dew_point + solar_radiation + rainfall + snowfall + season + holiday +
                is_weekend + rush_hour,
              data = train)
  pred <- predict(model, newdata = test)
  title_text <- "Regression: Actual vs. Predicted"
}
plot(test$rented_bike_count, pred,
     pch = 16, cex = 0.55, col = rgb(0, 121, 107, 85, maxColorValue = 255),
     main = title_text,
     xlab = "Actual rentals",
     ylab = "Predicted rentals")
abline(0, 1, col = pink, lwd = 3)
```

::: note
Predicted values closely follow actual rentals, but extreme demand periods remain harder to predict.
:::
::::

## Summary Findings for RQ1

::: slide-content
RQ1 asked: <br> **Which weather and time-related factors have the greatest influence on <br> hourly bike rental demand?**

### Key findings

-   **Temperature** is the strongest predictor of demand.
-   **Hour of day** is also highly influential, especially evening commute hours.
-   **Seasonality** matters: summer and autumn have higher demand than winter.
-   **Rainfall, snowfall, and humidity** are associated with lower demand.
-   Nonlinear modeling performs better than a simple linear model.
:::

## Business Implications

::: slide-content
For Seoul Bike Sharing operations managers:

-   Increase bike availability during high-demand evening hours.
-   Plan more resources during warmer seasons, especially summer and autumn.
-   Reduce expected allocation during rain, snow, or high-humidity periods.
-   Use predictive models to support staffing and maintenance scheduling.
-   Monitor weather forecasts as part of daily operational planning.
:::

## Limitations

::: slide-content
This analysis has several limitations:

-   The dataset is hourly and does not identify individual stations.
-   Station-level redistribution decisions cannot be directly modeled.
-   The model does not include external events, transit disruptions, or tourism patterns.
-   Random Forest improves prediction but is less interpretable than linear regression.
:::

## Conclusion

::: slide-content
The analysis shows that Seoul bike rental demand is strongly influenced by **temperature**, **hour of day**, **season**, and **weather conditions**.

The findings support a data-driven allocation strategy where managers adjust bike supply based on expected demand patterns rather than relying on fixed schedules.

### Final takeaway

Predictive demand analytics can improve service reliability, reduce operational inefficiency, and support better customer experience.
:::

# Thank You !

### Questions?

</div>
